### 2015-11-28 paper/1-introduction.tex
Rewrite/fixes in several sections. Initial section 6.

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a DEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
The dxexmachina kernel follows the design of the PythonPDEVS \cite{PythonPDEVS} kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The DEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on a (finite) representation of time. Dxexmachina kernels can operate on IEEE754 floating point time units or integral time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integral representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a secondary causality field. If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

>>>>

\subsection{Parallel computing}
%TODO parallel computing is necessary with more and more complex systems, blablabla (usual explanation)
Parallel execution of a PDEVS simulation can lower runtime and increase the bound on the state space, thereby enabling simulation of more complex systems in the same time-frame.
While the shared memory parallelism offered by most modern hardware does not increase the state space bounds, it can reduce the runtime and offers more direct communication and control between entities involved in synchronization compared to distributed simulation. \\

@@

\subsection{Solution} %for lack of a better word
%TODO give our solution, motivating the need for a different kernel, instead of modifying another kernel (like adevs or PythonPDEVS)
The usage of the DirectConnect\cite{SymbolicFlattening} algorithm makes reusing the adevs kernel hard.
The dxexmachina kernel follows the design of the PythonPDEVS \cite{PythonPDEVS} kernel where appropriate, but by the very nature of the implementation languages has to differ in key aspects (e.g. memory allocation strategy). 
The core aims of the project are to offer a deterministic simulation kernel where the simulation author is shielded as much as possible from the kernel implementation, without sacrificing performance. As in PyPDEVS, a model need be written only once for use in the different simulation kernels (with the exception of a non-trivial lookahead).\\
The tracing framework from PyPDEVS was ported to allow optional verification of simulations. 
\subsection{Time}
The PDEVS formalism has $\mathbb{R}$ as time base, but any implementation has to decide on an enumerable representation of time. Dxexmachina kernels can operate on IEEE754 floating point time units or integral time. In principle any type with well defined operators can be used as template parameter, but from a performance point of view a type fitting in a machine word offers obvious advantages.%ref to time paper.
An integral representation significantly reduces the possible range of the virtual time, but avoids approximation errors. Furthermore, the notion of $\epsilon$ as the absolute minimum between time points needs to be established, this is non-trivial for floating point.\\
The select() function, which imposes a sequential ordering between concurrent events, is implicit in this project by extending time representation with a causality field with a range at least as large as the maximum nr of models in the kernels ($2^{24}$ by default). If A and B are imminent at time t then $t[1]_a < t[1]_b \oplus t[1]_b < t[1]_a$, while $t[0]_a == t[0]_b == t[0]$. This avoids evaluating the select() function on all imminent models, while still maintaining the deterministic order of concurrent transitions.

### 2015-11-28 paper/2-background.tex
Rewrite/fixes in several sections. Initial section 6.

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of influencing models. A model can simulate up to (but not including) eit, and then has to wait until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other can be broken/avoided by a variety of
means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
Null message time is the timestamp a model is guaranteed to have past in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t is already sent.\\
In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
Conservative explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events. This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\
The performance is more sensitive to lookahead than it is to cycles in the (static) dependency graph.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but employs a roll-back mechanism to recover from those errors.
The most common mechanism is the Timewarp \cite{Jefferson:1985:VT:3916.3988} mechanism,
 whenever a kernel receives an event with timestamp in the kernel's past, the state of the kernel is reverted to that time. The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
Optimistic does not rely on any domain specific information, in contrast to conservative. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts worsened if the timestamps of the events match exactly. \\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the timepoint in the simulation that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
In any parallel implementation the GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

\subsection{C++11 Parallelism Features}
%TODO explain what is new in C++11 for parallelism and how it contributes to our cause
C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations. More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
Our kernels use a wide range of threading primitives and atomic operations. As an example, eot/eit/nulltime are exchanged not as messages but reads/writes to atomic fields shared by all kernels. This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages. % Antimessages ? More / Less here ?
Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform. %minus all the dependencies :)
% Reference C++11 standard (paying version or draft?)

>>>>

\subsection{Conservative Synchronization}
%TODO explain conservative synchronization, and how it uses locking to wait for continuing the simulation
Conservative synchronization is defined by the invariant that no model will advance in time before it has received all input from any influencing model. \\
This requires the concepts of eot (earliest output time) and eit (earliest input time) which define the timespan within which a model can safely advance. \\ The eit of any model is the minimum of all eot values of (directly) influencing models. A model can simulate up to (but not including) eit, then waits until that value is increased. An important disadvantage here is that the influenced-by relation is always defined at model(link) creation, not at runtime. A model that can influence another, but never does, can severely slow down the protocol.\\ 
Deadlock between models that influence each other and end up waiting on each other can be broken/avoided by a variety of
means, in this simulator the CMB \cite{Chandy:1981:ADS:358598.358613} null-message protocol is used. \\
In our implementaton null-time is the timestamp a model is guaranteed to have passed in simulation. More precisely, a null message of time t is a guarantee that any ouput with timestamp t-$\epsilon$ is already sent.\\
In general, the eot/eit/nulltime of a kernel is the mimimum of each of those values for all models in the kernel.\\
Conservative synchronization explicitly relies on information provided by the model creator in the form of lookahead, a relative timespan during which the model is insensitive to outside events. This can be non-trivial to calculate, a simulation writer will in general not be able to predict the exact lookahead of models involved in an experiment without having run the experiment.\\ Conservative kernels can operate if there exists a cyclic dependency between them, but at a quite severe performance penalty, as seen in section 4.
\subsection{Optimistic Synchronization}
%TODO explain optimistic synchronization, and how it uses rollbacks to undo causality errors
Optimistic synchronization allows causality errors to occur but recovers from those errors using a roll-back mechanism, the most common of which is Timewarp \cite{Jefferson:1985:VT:3916.3988}.
Whenever a kernel receives an event with a timestamp in the kernel's past, the state of the kernel (and all models) is reverted to that time. The gain in runtime this provides is offset by the increase in memory required to keep saved states and (sent) messages.
Optimistic does not rely on any domain specific information, in contrast to conservative. It is only sensitive to runtime use of connections, not the probability that they might occur. %reword connections ?
\\
If the (runtime) dependency graph contains a cycle, optimistic can suffer a series of cascading reverts. Without domain specific information the kernel assumes that any event will influence at least one model, but this can lead to an infinite loop of reverts in the worst case.\\
This effect can be lessened by lazy cancellation and/or lazy re-evaluation \cite{FujimotoBook}.
\subsection{Global Virtual Time}
To avoid exhausting memory in state/event saving, optimistic synchronization relies on the concept of global virtual time\cite{Jefferson:1985:VT:3916.3988}. In optimistic simulations, GVT is defined as the lowest timestamp of any unprocessed event. \\ Intuitively this is the simulation timepoint that is certain to be preserved, corresponding exactly with the simulation up to that time in a non-parallel implementation.\\
In conservative the minimum simulation time of all kernels is the GVT, or in terms of null messages: the least timestamp of any null message in transit.
The GVT calculation is vital to safely commit unrecoverable transactions such as IO (e.g. tracing), releasing memory, ... .

\subsection{C++11 Parallelism Features}
%TODO explain what is new in C++11 for parallelism and how it contributes to our cause
C++11 offers a wide range of portable synchronization primitives in the Standard Library, whereas in earlier versions one had to resort to non-portable (C) implementations. More importantly, C++11 is the first version of the standard that actually defines a multi-threaded abstract machine memory model in the language.
Our kernels use a wide range of threading primitives and atomic operations. As an example, eot/eit/nulltime are exchanged not as messages but reads/writes to atomic fields shared by all kernels. This avoids the otherwise unavoidable latency penalty by mixing simulation messages with synchronization messages, for an in-depth study see \cite{CPE:CPE3007}.
Most modern compilers support the full standard, allowing the kernels to be portable by default on any standard compliant platform.

### 2015-11-28 paper/3-features.tex
Rewrite/fixes in several sections. Initial section 6.

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS \cite{PythonPDEVS}  , and provides the following features: 
\begin{enumerate}
	\item Direct Connect \cite{SymbolicFlattening}
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable with adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PythonPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interaction 
	\item Distributed simulation
	\item Interactive control
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph vizualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
% weak scenarios for C, O and both ?
\subsubsection{Conservative}
% This needs better explaining, perhaps a central graph to reference to ?
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. Internal transitions are safe. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the eot of all influencing kernels. \\
For garbage collection (of sent messages) the GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $. %This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?.

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user.%Mention typed state here ?
The kernels use Mattern's %\cite{} %TODO find out in which paper this algorithm was described
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states/(anti)messages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks. 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with a reference count. Access of that reference count across threads is expensive. The string implementation also calls malloc/free which are quite expensive. \\
String are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance.
\\
\subsubsection{Smart pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the content (timestamp) directly into shared memory. This alleviates in part the potential delay/latency caused by interleaving simulation with null/eot messages in conservative implementations. Optimistic still sends anti-messages as pointers, since the complete header of the message is required at the receiver side to annihilate the original. 
\subsubsection{Schedulers}
PythonPDEVS has a wide range of scheduler for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depening on the nr of items to update. Items are only updated, never removed.

>>>>

%TODO explain that we are briefly going over the main features
\subsection{Based on PythonPDEVS}
%TODO say that it is based on PythonPDEVS
The simulator is based on PythonPDEVS \cite{PythonPDEVS}  , and provides the following features: 
\begin{enumerate}
	\item Direct Connect \cite{SymbolicFlattening}
	\item Dynamic Structured DEVS
	\item Termination function. If specified, a termination function is applied every simulation round to each model to test whether the simulation can terminate. Only available in single-threaded simulation.
	\item State/Message can have any payload type. Different message types can be used together within the same simulation.
	\item Tracing An asynchronous, thread safe and versatile tracing mechanism allows exact verification of the simulation.
	\item Optimistic and Conservative synchronization of PDEVS.
\end{enumerate}
The implementation tries to adhere to the C++ principle that you don't pay performance-wise for what you don't use. For this reason, the support for a termination function for the multi-threaded kernel was abandoned, as it is non-trivial to implement and had a non-negligible impact on the runtime, even when not in use.\\
The tracing is not comparable with adevs's listener interface. To be usable in optimistic simulation, the tracing of the simulation has to be reversible and only be committed at GVT points. Furthermore, the framework itself has to be threadsafe and deterministic so that a simulation will always produce the exact same output.
%TODO Decide whether we should list the functionality that is _not_ implemented
The following features from PythonPDEVS are not present
\begin{enumerate}
	\item Activity tracking and relocation
	\item Serialization
	\item Interactive control. 
	\item Distributed simulation
\end{enumerate}
Serialization in this context is the ability to save/load a complete simulation to disk, not the state saving mechanisms required for TimeWarp.\\ State saving has no impact in a single threaded or conservative kernel.\\
Model allocation is done by a derivable allocator object which the user can implement to arrange a more ideal (domain-specific) allocation. If this is omitted, a default (non-activity-aware) allocation stripes the models over the simulation kernels.\\
Debugging tools such as a logger and a graph visualizer are included which can track activity with respect to allocation for later study, but not online/dynamic as is possible in PyPDEVS.\\
While PythonPDEVS can be controlled from within a Python script and adevs has a Java interface, our implementation does not have any bindings to other languages.

%TODO ... and thus has most features from PythonPDEVS, like Dynamic Structure, termination condition, direct connection, ...
%TODO (size: take as much size as you want, depending on how much is shared with PythonPDEVS)
%TODO (maybe: note that models are not compatible)
\subsection{Different Synchronization protocols}
%TODO explain rationale: why different synchronization protocols
% weak scenarios for C, O and both ?
\subsubsection{Conservative}
% This needs better explaining, perhaps a central graph to reference to ?
A conservative kernel will determine which kernels it is influenced by. This information is constructed from the incoming connections on all hosted models. The process is only 1 link deep, since an influenced kernel will in turn be blocked by others deeper in the graph.\\
A model should provide a lookahead function which returns, relative from the current time, the timespan during which the model cannot change state due to an external event. This information is collected for all models hosted on the kernel, and the minimum is set as the lookahead of that kernel. \\
The kernel will calculate its earliest output time and write this value in shared memory. The eit of the kernel is then set as the minimal eot of all influencing kernels. \\
For garbage collection (of sent messages) the LBTS/GVT is calculated as $\min_{\forall i \in \textup{influencors}}( \textup{nulltime}[i]\ )  - \epsilon $. %This is one of the operations that can benefit from using a relaxed memory ordering. % Using it for read atm, check before finalizing and do we need to mention this ?.

\subsubsection{Optimistic}
The optimistic kernel requires from the hosted model only that copying the state is well-defined, which is provided in the base State class for the user. 
The kernels use Mattern's \cite{mattern}
GVT algorithm with a maximum of 2 rounds per iteration to determine a GVT. This process runs asynchronously from the simulation itself. Once found, the controlling thread informs all kernels of the new value, which they can use to execute garbage collection of old states/(anti)messages.

@@

\subsection{Performance Improvements}
%TODO briefly discuss some of the additional changes you implemented
%TODO such as memory pools, memory allocators, profiling-guided-optimizations, ...
Continuous profiling of the kernels in several benchmarks highlighted the following key bottlenecks: 
\subsubsection{Heap}
A kernel never sends a complete object to another kernel, only a pointer to the object. This avoids a possibly expensive copy of the payload, but at the cost of allocation overhead.\\
This cost becomes prohibitively expensive in highly connected models, so to reduce that overhead we use thread\_local memory pools for states and messages, and optionally replace the system malloc with calls to tcmalloc\cite{tcmalloc}. In this way allocating threads do not block each other, and in a single threaded kernel we can leverage arena-style pools if desired.\\
This changed the ownership semantics of several objects in a non-trivial way, since the thread that creates an object has to destroy it (iff it can prove it is no longer used). Experiments with synchronized pools proved slower than malloc/free. \\
Initially the kernels used strings as identifiers, as is done in PyPDEVS, profiling quickly indicated this to be a performance bottleneck. C++ strings are heap allocated variable sized objects with an atomic reference count. Access of that reference count across threads is expensive, as are the calls to malloc/free the string implementation makes to create/destroy new object, or copy existing.\\
Strings are more intuitive to work with from a user's standpoint, so as a compromise the user can reference models/ports by string name (usually when constructing the model). Once simulation starts all objects use integral identifiers for performance. This also increased usage of the constexpr feature of C++11 in, amongst others, timestamps and message headers.
\\
\subsubsection{Smart pointers}
While an important C++11 feature in general, our initial usage of smart pointers for some types of objects was misplaced. Used across threads the reference counting becomes prohibitively expensive, and the (de)allocating caused significant contention between threads. Models are still held by a smart pointer, as is a kernel, but a message is a raw pointer to compacted memory. 
\subsubsection{Locking}
Locking between kernels uses mostly atomic operations, where we can occasionally leverage memory orderings to only pay for synchronization when we need it. Messages are exchanged via a shared set of queues each with a dedicated lock.\\
On a higher level, we avoid the sending of synchronization messages entirely by writing the timestamp directly into shared memory.\\ Sending of antimessages is fairly cheap in our implementation, since only the modified pointer to the original message is sent to the receiving kernel.
\subsubsection{Schedulers}
PythonPDEVS has a wide range of schedulers for the user to choose from, with performance of each depending on the simulation type. Profiling showed in our case that, for a c++ implementation, the heap implementation used in adevs was faster than any of the schedulers we had tested before. Unlike most node based heaps, this scheduler uses a fixed size array where a heap is rebuilt or modified in place depending on the nr of items to update. Items are only updated, never removed.

### 2015-11-28 paper/4-performance.tex
Rewrite/fixes in several sections. Initial section 6.

Interconnect is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models. It highlights in sequential simulaton the cost of heap allocation (messages) in dxexmachina. As the benchmark increases the model count, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t. adevs. This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains a significant effect.

@@

Since dx's conservative passes messages by pointer, it needs a GVT implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
Optimistic needs a far more complex GVT/garbage collection algorithm which explains in part the massive amount of memory used. Devstone (flattened) is allocated in a chain, this means that the leafs in the dependency graph will do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
% split: dstone/pqueue are dag's
\subsubsection{Devstone}
% Initial.
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic this can cause more reverts since the kernels will start to drift faster as the modelcount increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for n=dxw models, it takes timeadvance()*n transitions to activate the last model in the chain. For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
\subsubsection{PHold}
%Phold
In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of subnodes. The parameter R determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the nulltime of N-1 kernels, resulting in O($N^2$) polling behaviour. This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message. In a non-cyclic simulation with a non-trivial lookahead, that choice however does pay off.\\
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. More seriously, a revert is very expensive in PHold due to our usages of C++11 's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between the kernels increases fast, increasing the likelyhood of more reverts.

\subsubsection{Interconnect}
In Interconnect % no paper ? from PyPDevs originally ? 
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as in PHold.\\
Our optimistic implementation does not complete an instance of this benchmark. The kernels get stuck in an infinite cascade of reverts. If kernel A reverts, it will send antimessages to all others who in turn revert and send antimessages to all others (and A itself again). Support for lazy cancellation could potentially undo this anti-pattern.
\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ The priority messages defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages. 
\input pqueue.tex
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create, once the random nr generation is done the kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.

>>>>

Interconnect \cite{van2013research} is a benchmark where all models broadcast, creating a complete graph in terms of dependencies between models. It highlights in sequential simulation the cost of heap allocation (messages) in dxexmachina. As the model count increases, we see the expected quadratic increase in runtime in both kernels, but an increasing penalty for dxexmachina w.r.t. adevs. This effect is due to the heap allocation of the messages by dxexmachina, even though this is minimized using memory pools it remains significant.

@@

Since conservative passes messages by pointer, it needs a GVT/LBTS implementation to organise garbage collection, this inevitable delay explains the higher memory usage w.r.t adevs.\\
Optimistic needs a more complex GVT/garbage collection algorithm plus the differences in LP virtual times are far larger, which explains the heavier memory usage. Devstone (flattened) is allocated in a chain, this means that the leafs in the dependency graph will do a lot of unnecessary simulation before having a revert, leading to quite severe memory pressure. Unlike conservative and sequential execution, memory usage in optimistic varies greatly depending on scheduling of kernel threads and drifting between kernels. 

@@

\subsection{Parallel Simulation}
%TODO explain that we will now compare different synchronization protocols
%TODO compare different synchronization protocols with adevs' conservative, showing that we are faster (or at least, not that much slower) than adevs
% split: dstone/pqueue are dag's
\subsubsection{Devstone}
% Initial.
The flattened models are allocated to kernels by giving each kernel a distinct section of the chain, resulting in a low ratio of inter-kernel to intra-kernel messages. For optimistic this can cause more reverts since the kernels will start to drift faster as the modelcount increases. Furthermore, optimistic is quite sensitive to an increase in kernels, since the delay before a revert propagates increases. Of note as well is the warm-up time this benchmark requires, for n=dxw models, it takes timeadvance()*n transitions to activate the last model in the chain. For parallel this can be reduced to $\frac{ta()*n}{kernels}$ before the last kernel becomes active.
\subsubsection{PHold}
%Phold
In Phold \cite{PHOLD} Allocation is specified in the benchmark itself, each kernel manages a single node with a constant set of sub-nodes. The parameter R determines the percentage of remote destination models.\\
The dynamic dependency graph is a very sparse version of the static dependency graph, penalizing conservative. Lookahead is $\epsilon$, so conservative spends most of its time crawling in steps of $\epsilon$. Since the dependency graph between kernels is a complete graph, this is not a simulation that scales in our implementation. For N kernels, each kernel has to query the null-time of N-1 kernels, resulting in O($N^2$) polling behaviour. This benchmark highlights therefore the price that we pay for sharing those values instead of sending an actual null message. In a non-cyclic simulation with a non-trivial lookahead (e.g. devstone), that choice however does pay off.\\
Optimistic suffers little from the above problems, however due to the high interconnectivity a cascading revert is still possible. More seriously, a revert is very expensive in PHold due to our usage of C++11's random nr generators. The cost of a revert is dominated by the recalculation of destination models, not in allocating/deallocating states/messages. Again, this could be significantly reduced using lazy-cancellation. Once a revert happens the drift between the kernels increases fast, increasing the likelihood of more reverts.

\subsubsection{Interconnect}
In Interconnect
the set of atomic models form a complete graph (w.r.t. connections), each model broadcasts messages to the entire set. \\
Allocation is irrelevant, the resulting dependency graph between kernels remains  a complete graph. The runtime dependency graph is almost immediately equal to the static dependency graph.
Conservative still faces the same issues as in PHold, with the key difference that for a fixed time advance lookahead is equal to the timespan between transitions. The scaling issue is identical as in PHold.\\
Our optimistic implementation does not complete an instance of this benchmark. The kernels get stuck in an infinite cascade of reverts. If kernel A reverts, it will send anti-messages to all others who in turn revert and send anti-messages to all others (and A itself again). Support for lazy cancellation could potentially undo this anti-pattern.
\subsubsection{Priority network model}
%TODO show a single plot which varies a parameter that changes the ideal synchronization protocol
The priority benchmark is composed of a single server generating a stream of $0<=m<=n$ messages at fixed time intervals, interleaved with a probability p for a priority message, to n receivers. \\ This defaults lookahead for the receivers to $\epsilon$ but this time there is no scaling effect, nor are there cycles in the dependency graph. This model therefore highlights the basic strengths/weaknesses of both synchronization protocols. Receiving models are allocated on another kernel than the server, and have a internal transition so will not wait for the incoming messages. 
\input pqueue.tex
A key difference here with the other benchmarks is that a state (in the Receiver instances) is very cheap to copy/create, once the random nr generation is done the kernel holding the server will never revert since it is a source in the dependency graph. Optimistic will therefore not suffer the same performance hit in recreating states as it does in PHold.

### 2015-11-28 paper/6-conclusions.tex
Rewrite/fixes in several sections. Initial section 6.

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))

>>>>

%TODO conclude the paper, saying that our new simulation tool offers conservative and optimistic synchronization, resulting in good performance, which is even better than adevs
% needs some extending. Maybe clarify that we actually do single core PDEVS and not
% pure classic DEVS.
Both synchronization algorithms offer good performance in differing simulations, in some simulation our kernels outperform adevs whereas in others we can still improve. 
The optimistic implementation needs to be extended with lazy evaluation/cancellation to function in cyclic simulations.
\subsection{Future work}
%TODO future work: more things that we might want to implement, and probably something about automatically switching between synchronization protocols at runtime (and reference my activity papers ;))
\subsubsection{Activity}
As shown in \cite{PythonPDEVS_ACTIMS} activity and allocation of models across kernels is a key aspect in achieving high performance in any parallel implementation. Allocating models so that there are no dependency cycles between their containing kernels is a first step, but not always possible. For optimistic one can use re-allocation to break (runtime dependency) cycles or perform load balancing. If kernels are unevenly balanced they will begin to drift fast, causing increasingly more reverts. 
\subsubsection{Hybrid}
The optimistic implementation could use (null/eot/eit) from conservative to detect and/or reduce the cost of reverts without completely stalling on influencing kernels.
Conservative kernels could be extended with runtime information about influencing models, if one can guarantee a static dependency is not used for a fixed time-span, this dependency can be removed for that period of (virtual) time. \\
Ultimately the simulation could switch at runtime between protocols based on the information provided by activity tracking.
